---
title: "MALASPINA SRF | 16S miTags | Bacteria"
subtitle: "Dataset: mapping_SILVA_plus_PhytoRef_plus_CyanoDB_NR99.txt"
date: "02.02.2015"
oumput:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    highlight: zenburn
    toc: yes
---

<!--- INITIALIZATION
```{r, echo=FALSE}
#error hook to kill knitr in case of errors
library(knitr)
knit_hooks$set(error = function(x, options) stop(x))
opts_chunk$set(cache=TRUE, autodep=TRUE)
```
--->

# 1) Data overview

Samples with less than 167 reads are removed: 

``` {r load_data, echo=FALSE, message=FALSE}
#setwd("/genwork/lrubinat")
setwd("/home/laura/Documents/TFM/genwork/data_analysis/MP_SRF_miTags_NR99/MP_SRF_miTags_NR99_bacteria/")

#read data 
mp_srf_16SmiTags_b <- read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/MP_SRF_miTags_NR99_bacteria.txt", head=TRUE, fill=TRUE)

#table dimensions and format before setting column names
dim(mp_srf_16SmiTags_b) # 44068   274
mp_srf_16SmiTags_b[1:5,1:5]

#row names = OTU name (option A)
row.names(mp_srf_16SmiTags_b)<-mp_srf_16SmiTags_b[,1]

mp_srf_16SmiTags_b<-mp_srf_16SmiTags_b[,-1]
mp_srf_16SmiTags_b[is.na(mp_srf_16SmiTags_b)]<-0

dim(mp_srf_16SmiTags_b)
mp_srf_16SmiTags_b[1:5,1:5]

colnames(mp_srf_16SmiTags_b)
#remove samples with less than 167 reads
mp_srf_16SmiTags_b_min<-mp_srf_16SmiTags_b[,colSums(mp_srf_16SmiTags_b) > 167]
amplicons_per_sample_mp_srf_16SmiTags_b<-colSums(mp_srf_16SmiTags_b)
amplicons_per_sample_mp_srf_16SmiTags_b[which(colSums(mp_srf_16SmiTags_b)>40000)]



#remove samples with less than 167 reads in the "mp_srf_16S_miTags_99tb_chlp_cyanob" and "mp_srf_16S_miTags_99tb_euk" datasets (so that we can compare the relative abundance of 16S and 18S OTUs considering the same samples)
mp_srf_16SmiTags_b_v2<-subset(mp_srf_16SmiTags_b_min167, select=-c(X124_DCM_0d2.0d45, X124_SUR_0d2.0d45, X125_SUR_0d2.0d45, X36_SUR_0d1.0d2, X37_OMZ_0d2.1d6, X38_OMZ_0d2.1d6, X38_SUR_0d1.0d2, X39_OMZ_0d2.1d6, X39_SUR_0d1.0d2, X41_SUR_0d1.0d2, X7_SUR_0d2.1d6, X93_SUR_0d2.3, X123_DCM_0d1.0d2, X125_DCM_0d1.0d2, MP0317, MP0320, MP0322, MP0536, MP0784, MP0882, MP0884, MP0886, MP1164, MP1166, MP1178, MP1337, MP1413, MP1415, MP1419, MP1523, MP1525, MP1599, MP1639, MP1677, MP1678, MP1679, MP1681, MP1682, MP1849, MP1851, MP1853, MP1879, MP2000, MP2043, MP2235, MP2237, MP2658, MP2817, MP427, MP46, MP464, MP468, MP692))
```

Table dimensions and content outline:

```{r starting_dataset, echo=FALSE}
dim(mp_srf_16SmiTags_b_v2)
mp_srf_16SmiTags_b_v2[1:5,1:5]
```

Minimum number of reads per station:

```{r reads_per_sample_overview1, echo=1}
min(colSums(mp_srf_16SmiTags_b_v2)) 
```

Maximum number of reads per station:

```{r reads_per_sample_overview2, echo=1}
max(colSums(mp_srf_16SmiTags_b_v2)) 
```

Identification of station with higher number of reads:

```{r reads_per_sample_overview3, echo=TRUE}
amplicons_per_sample_mp_srf_16SmiTags_c<-colSums(mp_srf_16SmiTags_b_v2)
amplicons_per_sample_mp_srf_16SmiTags_c[which(colSums(mp_srf_16SmiTags_b_v2)>160000)]
```

Overall reads per sample:

``` {r reads_per_sample_overview4, echo=FALSE}
plot(sort(colSums(mp_srf_16SmiTags_b_v2)), pch=19, xlab="sample", ylab="reads per sample", cex=0.9)
```

# 2) Normalization to 167 reads per sample

Let's normalize the original dataset by randomly subsampling 167 reads in each station:

``` {r species_richness_rarefaction1, echo=TRUE}
library(vegan)
mp_srf_16SmiTags_b_v2_t<-t(mp_srf_16SmiTags_b_v2)
mp_16mt_b_ss167<-rrarefy(mp_srf_16SmiTags_b_v2_t, 167)
```

The normalized table shows the following dimensions and format:

```{r species_richness_rarefaction2, echo=FALSE}
dim(mp_16mt_b_ss167)
mp_16mt_b_ss167[1:5,1:5]
```

Its content fits with the expected normalization values (167 reads per station):

``` {r species_richness_rarefaction3, echo=TRUE}
rowSums(mp_16mt_b_ss167)
```

Let's check out how many OTUs don't appear in the new table:

```{r species_richness_rarefaction4, echo=1:5}
length(which(colSums(mp_16mt_b_ss167)==0)) 
```

There are 37653 OTUs that don't show any occurrence in the normalized data. Let's remove them from the table and take a look at its final dimensions:

```{r species_richness_rarefaction5, echo=1:3}
mp_16mt_b_ss167_no_cero<-mp_16mt_b_ss167[,-(which(colSums(mp_16mt_b_ss167)==0))]
dim(mp_16mt_b_ss167_no_cero)

#The final dimensions of the normalized table are  220 2101.
#6416+37653 = 44069
```

Datasets summary:

mp_srf_16SmiTags_b_v2_t --> 220 44069

mp_16mt_b_ss167_no_cero --> 220 6416


# 3) General community analysis

## 3.1) Richness and evenness (Shannon index)

```{r shannon_index1, echo=FALSE}
mp_16mt_b_ss167_div <- diversity(mp_16mt_b_ss167_no_cero, index="shannon")
```

Most of the samples take Shannon Index values between 4.7 and 4.9:

```{r shannon_index2, echo=FALSE}
boxplot(mp_16mt_b_ss167_div, pch=19, main="Shannon's index of diversity")
plot(sort(mp_16mt_b_ss167_div), pch=19, main="Shannon's index of diversity")
```

## 3.2) Richness: OTU number

```{r richness_otu_no1, echo=FALSE}
OTUs_per_sample_mp_16mt_b_ss167<-specnumber(mp_16mt_b_ss167_no_cero)
```

Lowest number of OTUs per sample:

```{r richness_otu_no2, echo=FALSE}
min(OTUs_per_sample_mp_16mt_b_ss167)
```

Maximum number of OTUs per sample:

```{r richness_otu_no3, echo=FALSE}
max(OTUs_per_sample_mp_16mt_b_ss167)
```

In most of the samples, we can identify between 130 and 145 OTUs:

```{r richness_otu_no4, echo=TRUE}
plot(sort(OTUs_per_sample_mp_16mt_b_ss167), pch=19)
boxplot(OTUs_per_sample_mp_16mt_b_ss167, pch=19)
```

## 3.3) Index of evenness

### 3.3.1) Pielou's index

```{r pielou_index_of_b_venness1, echo=TRUE}
pielou_b_venness_mp_16mt_b_ss167 <- mp_16mt_b_ss167_div/log(OTUs_per_sample_mp_16mt_b_ss167)
```

The Pielou index (constrained between 0 and 1) takes values closer to 1 as the variation of species proportion in a sample decreases. Most of our samples get valuesa around 0.95, meaning that the numerical composition of different OTUs within our samples is so similar:

```{r pielou_index_of_b_venness2, echo=TRUE}
plot(sort(pielou_b_venness_mp_16mt_b_ss167), pch=19)
boxplot(pielou_b_venness_mp_16mt_b_ss167, pch=19)
```

The OTU_3009, with 413 reads, is the most abundant in the overall dataset:

```{r OTUs_overall_abundance, echo=TRUE}
head(sort(colSums(mp_16mt_b_ss167_no_cero), decreasing=T), n=10L)
```

Most of the OTUs show few occurrences:

```{r OTUs_overall_abundance2, echo=TRUE}
plot(log(sort(colSums(mp_16mt_b_ss167_no_cero), decreasing=T)), pch=19)
```

<!---
### 3.3.2) Sads
library(sads)
?sads
--->

## 3.4) Abundance Models

<!---
### 3.4.1) Rank-Abundance or Dominance/Diversity Model ("radfit")

To be solved!!

```{r radfit, echo=FALSE}
#?radfit
#otu_tb18_t[1:5,1:5]

mp_srf_16SmiTags_b_v2_t_radfit<-radfit(colSums(mp_srf_16SmiTags_b_v2_t))
plot(mp_srf_16SmiTags_b_v2_t_radfit)
```
--->

### 3.4.2) Preston's Lognormal Model

According to Preston's lognormal model fit into species frequencies groups, we're missing ~80813 species:

```{r preston_model1, echo=T}
mp_16mt_b_ss167_prestonfit<-prestonfit(colSums(mp_srf_16SmiTags_b_v2_t))
plot(mp_16mt_b_ss167_prestonfit, main="Pooled species")

veiledspec(mp_16mt_b_ss167_prestonfit)
```

When computing Preston's lognormal model fit without pooling data into groups, we seem to miss ~57631 species:

```{r preston_model2, echo=4}
mp_16mt_b_ss167_dist_all<-prestondistr(colSums(mp_srf_16SmiTags_b_v2_t))
plot(mp_16mt_b_ss167_prestonfit, main="All malaspina")
lines(mp_16mt_b_ss167_dist_all, line.col="blue3")

veiledspec(mp_16mt_b_ss167_dist_all)
```

<!---
## 3.5) Rarefaction curve

(To be computed)

```{r rarefraction_curve, echo=TRUE}
#?rarecurve

#str(colSums(otu_tb18_t))

#otus_tb18_colsums<-colSums(otu_tb18_t)

#str(otus_tb18_colsums)
#otu_tb18_colsums<-as.matrix(otu_tb18_colsums)

#otu_tb18_colsums<-t(otu_tb18_colsums)

#otu_tb18_colsums[,1:3]

#rarecurve(otu_tb18_colsums, step = 1, 50000, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
#rarecurve(colSums(otu_tb18_t), step = 1, 50000, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
```
--->

## 3.6) Beta diversity

### 3.6.1) Dissimilarity matrix using Bray-Curtis index:

The Bray-Curtis dissimilarity, constrained between 0 (minimum distance) and 1 (highest dissimilarity) allows us to quantify the differences between samples according to the composition and relative abundance of their OTUs. In our dataset, most of the samples pairs take dissimilarity values between 0.85 and 0.95, meaning that their composition is highly different.

```{r beta_div1, echo=FALSE}
#?vegdist
mp_16mt_b_ss167_no_cero.bray<-vegdist(mp_16mt_b_ss167_no_cero, method="bray")
boxplot(mp_16mt_b_ss167_no_cero.bray, main="Bray-Curtis dissimilarity matrix")
```

### 3.6.2) Hierarchical clustering

We do observe distinguished clusters; we'll need to implement further analysis to determine if the observed groups are meaningful.

(To be done: add Longhurst provinces and sampling depth information)

```{r beta_div2, echo=FALSE}
#UPGMA
mp_16mt_b_ss167_no_cero.upgma<-hclust(mp_16mt_b_ss167_no_cero.bray, "average")
plot(mp_16mt_b_ss167_no_cero.upgma, cex=.35, main="Samples Hierarchical Clustering")
```

### 3.6.3) Non-metric multidimensional scaling

In general terms, we can distinguish a big cluster in the middle of the chart and two weak groups clustering in the middle-right side of the edge.

The stress parameter takes a value below 0.3, meaning that the plot is valid. 

```{r monoNMDS, echo=F}
#NMDS
mp_16mt_b_ss167_no_cero.nmds<-monoMDS(mp_16mt_b_ss167_no_cero.bray)
mp_16mt_b_ss167_no_cero.nmds
plot(mp_16mt_b_ss167_no_cero.nmds, main="monoMDs method")
```

When implementing a most robut function for computing NMDS plots, the result is quiet the same:

```{r metaNMDS, echo=F}
mp_16mt_b_ss167_no_cero.meta_nmds<-metaMDS(mp_16mt_b_ss167_no_cero.bray)
plot(mp_16mt_b_ss167_no_cero.meta_nmds, main="metaMDS method")
```

# 4) Geographical analysis

```{r load_geo_data, echo=F, results="hide", message=F}
#load geographical ubication of stations and sort according to otu_tb16 stations sequence.
mp_geo_16S_b_ss167<-read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/geo_ubication_MP_SRF.txt", sep="\t", header=T)

row.names(mp_geo_16S_b_ss167)<-mp_geo_16S_b_ss167[,1]
mp_geo_16S_b_ss167_sorted<-mp_geo_16S_b_ss167[row.names(mp_16mt_b_ss167_no_cero),]

dim(mp_geo_16S_b_ss167)
mp_geo_16S_b_ss167[1:5,1:4]
mp_16mt_b_ss167_no_cero[1:5,1:5]

#read lat-long in decimal degrees and translate into distance in km.
library(fossil)

#select only columns containing info about station, latitude and longitude.
mp_geo_16S_b_ss167_v2<-create.lats(mp_geo_16S_b_ss167, loc="sample", long="long", lat="lat")
head(mp_geo_16S_b_ss167)

#create a distance matrix (lower triangle) between a list of points.
geo_distances_mp_16S_b_ss167<-earth.dist(mp_geo_16S_b_ss167_v2, dist = TRUE)
head(geo_distances_mp_16S_b_ss167)
dim(geo_distances_mp_16S_b_ss167)

geo_distances_mp_16S_b_ss167<-as.matrix(geo_distances_mp_16S_b_ss167)
dim(geo_distances_mp_16S_b_ss167)

#geo distances dataset ready to use "geo_distances_MP_b_uks"
```

Working datasets:

1) Community matrix: mp_16mt_b_ss167_no_cero

```{r working_datasets1, echo=T}
dim(mp_16mt_b_ss167_no_cero)
mp_16mt_b_ss167_no_cero[1:5, 1:5]
```

2) Community Bray-Curtis: otu_tb16_cc_t_ss5000_no_cero.bray

```{r working_datasets2, echo=2}
#dim(otu_tb16_t_ss5000_no_cero.bray)
mp_16mt_b_ss167_no_cero.bray<-as.matrix(mp_16mt_b_ss167_no_cero.bray)
```

3) Stations distances in km: geo_distances_mp_16S_b_ss167

```{r working_datasets3, echo=T}
dim(geo_distances_mp_16S_b_ss167)
```

Communities quickly change their composition across geographical distances:

```{r working_datasets4, echo=T}
plot(geo_distances_mp_16S_b_ss167, mp_16mt_b_ss167_no_cero.bray, pch=19, cex=0.4, xlab="Geopgraphical distances", ylab="Bray-Curtis dissimilarities")
```

## 4.1) Mantel correlograms

The Mantel Test statistic is significant, but it is so low, meaning that the correlation between samples composition and geographical distance is very weak.

```{r mantel_correlogram1, echo=T}
mantel(geo_distances_mp_16S_b_ss167, mp_16mt_b_ss167_no_cero.bray)
```

Maximum distance between samples:

```{r mantel_correlogram2, echo=F}
max(geo_distances_mp_16S_b_ss167)
```

Minimum distance between samples:

```{r mantel_correlogram3, echo=F}
min(geo_distances_mp_16S_b_ss167)
```

Correlograms:

```{r mantel_correlogram4, echo=T}

mp_16S_b_ss167_mantel_correl_by_1000km<-mantel.correlog(mp_16mt_b_ss167_no_cero.bray, D.geo=geo_distances_mp_16S_b_ss167, break.pts=seq(0,20000, by=1000))
plot(mp_16S_b_ss167_mantel_correl_by_1000km)

mp_16S_b_ss167_mantel_correl_by_100km<-mantel.correlog(mp_16mt_b_ss167_no_cero.bray, D.geo=geo_distances_mp_16S_b_ss167, break.pts=seq(0,20000, by=100))
plot(mp_16S_b_ss167_mantel_correl_by_100km)
```

# 5) Abundance & occurence

```{r OTUs_mean_relative_abund, echo=F, results="hide"}
mp_16mt_b_ss167_no_cero[1:5,1:5]
mp_16mt_b_ss167_no_cero_t<-t(mp_16mt_b_ss167_no_cero)

colSums(mp_16mt_b_ss167_no_cero_t)

#local abundance percentage
mp_16mt_b_ss167_no_cero_t.rabund<-mp_16mt_b_ss167_no_cero_t/167

colSums(mp_16mt_b_ss167_no_cero_t.rabund)
mp_16mt_b_ss167_no_cero_t.rabund[1:5,1:5]

#OTUs mean relative abundance
mp_16mt_b_ss167_no_cero_t.rabund_means<-rowMeans(mp_16mt_b_ss167_no_cero_t.rabund) 
mp_16mt_b_ss167_no_cero_t.rabund_means<-?(mp_16mt_b_ss167_no_cero_t.rabund_means)

head(mp_16mt_b_ss167_no_cero_t.rabund_means)
```

```{r OTUs_occurence, echo=F, results='hide'}
mp_16mt_b_ss167_no_cero_t.rabund.occur<-mp_16mt_b_ss167_no_cero_t.rabund
mp_16mt_b_ss167_no_cero_t.rabund.occur[mp_16mt_b_ss167_no_cero_t.rabund.occur>0]<-1
mp_16mt_b_ss167_no_cero_t.rabund.occur[1:5,1:5] ### presence - absence table

#percentage of occurence in overall stations
mp_16mt_b_ss167_no_cero_t.rabund_means.occurence_perc<-as.data.frame(100*(rowSums(mp_16mt_b_ss167_no_cero_t.rabund.occur)/220))

str(mp_16mt_b_ss167_no_cero_t.rabund_means.occurence_perc)
```

```{r merge_rabund_peroccur, echo=F, results='hide'}
mp_16mt_b_ss167_rabund_percoccur<-merge(mp_16mt_b_ss167_no_cero_t.rabund_means, mp_16mt_b_ss167_no_cero_t.rabund_means.occurence_perc, by="row.names")

colnames(mp_16mt_b_ss167_rabund_percoccur)<-c("OTUs","mean_rabund","perc_occur")
mp_16mt_b_ss167_rabund_percoccur[1:5,]

row.names(mp_16mt_b_ss167_rabund_percoccur)<-mp_16mt_b_ss167_rabund_percoccur[,1]
mp_16mt_b_ss167_rabund_percoccur<-mp_16mt_b_ss167_rabund_percoccur[,-1]
mp_16mt_b_ss167_rabund_percoccur[1:5,]
```

In the following plot, we can appreciate the OTUs distribution according to their percentage of occurence and relative abundance. The blue line restricts regionally abundant OTUs (> 0.1%).

```{r abund_vs_occurence_table, echo=F}
plot(mp_16mt_b_ss167_rabund_percoccur$mean_rabund,mp_16mt_b_ss167_rabund_percoccur$perc_occur, log="x", pch=19, cex=0.8, xlab="Mean relative abundance", ylab="Percentage of occurence")
abline(h=80, col="red") #occurence higher than 80%
abline(v=0.00001, col="green") #rare OTUs
abline(v=0.001, col="blue") #cosmopolitan OTUs

#Conventional limits:
#Regionally rare     = 0.00001
#Regionally abundant = 0.001
```

Regionally abundant OTUs (relative abundance over 0.1%):

```{r abundant_OTUs, echo=7}
#regionally abundant
mp_16mt_b_ss167_abundant<-mp_16mt_b_ss167_rabund_percoccur[mp_16mt_b_ss167_rabund_percoccur$mean_rabund > 0.001,]

mp_16mt_b_ss167_abundant_sorted<-mp_16mt_b_ss167_abundant[order(mp_16mt_b_ss167_abundant$mean_rabund, mp_16mt_b_ss167_abundant$perc_occur, decreasing = T), c(1,2)]

mp_16mt_b_ss167_abundant_sorted
dim(mp_16mt_b_ss167_abundant_sorted)
```

Proportion of regionally abundant OTUs (%):

```{r abundant_OTUs2, echo=F}
#there are 84 regionally abundant OTUs.
(137/6416)*100 # = 6.22% of the OTUs are regionally abundant

#length(row.names(otu_tb16_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund > 0.001,])) # 83 OTUs
#row.names(otu_tb16_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund > 0.001,])
```

Cosmopolitan OTUs (abundance over 0.1% and occurence in more than 80% of samples):

```{r select_cosmopolitan, echo=6}
mp_16mt_b_ss167_rabund_cosm<-mp_16mt_b_ss167_rabund_percoccur[mp_16mt_b_ss167_rabund_percoccur$mean_rabund > 0.001,]
mp_16mt_b_ss167_rabund_poccur_cosm<-mp_16mt_b_ss167_rabund_cosm[mp_16mt_b_ss167_rabund_cosm$perc_occur > 80,]
mp_16mt_b_ss167_cosmop_sorted<-mp_16mt_b_ss167_rabund_poccur_cosm[order(mp_16mt_b_ss167_rabund_poccur_cosm$perc_occur, mp_16mt_b_ss167_rabund_poccur_cosm$mean_rabund, decreasing = T), c(1,2)]

mp_16mt_b_ss167_cosmop_sorted
dim(mp_16mt_b_ss167_cosmop_sorted)
```

Proportion of cosmopolitan OTUs (%):

```{r percentage_cosmopolitan, echo=F}
(0/6416)*100
```

Number and proportion (%) of rare OTUs:

```{r rare_OTUs, echo=1}
dim(mp_16mt_b_ss167_rabund_percoccur[mp_16mt_b_ss167_rabund_percoccur$mean_rabund < 0.00001 & mp_16mt_b_ss167_rabund_percoccur$mean_rabund >0,])
 
(0/6416)*100 # = 29.22% of the OTUs are regionally rare
```

We can't identify any instance of cosmopolitan or rare OTUs (!).

<!---
```{r otu_col_chech, echo = T}
dim(mp_16mt_b_ss167_rabund_percoccur)

#rare OTUs:
# length(row.names(otu_tb16_cc_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund < 0.00001 & otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund >0 ,])) # 325 OTUs
```
--->