---
title: "MALASPINA SRF | 10 miTags samples - ss | Bacteria"
subtitle: "Dataset: mapping_SILVA_plus_PhytoRef_plus_CyanoDB_NR99.txt"
date: "02.02.2015"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    highlight: zenburn
    toc: yes
---

<!--- INITIALIZATION
```{r, echo=FALSE}
#error hook to kill knitr in case of errors
library(knitr)
knit_hooks$set(error = function(x, options) stop(x))
opts_chunk$set(cache=TRUE, autodep=TRUE)
```
--->

# 1) Data overview

Samples with less than 13674 reads are removed: 

``` {r load_data, echo=FALSE, message=FALSE}
#setwd("/genwork/lrubinat")
setwd("/home/laura/Documents/TFM/genwork/data_analysis/MP_SRF_miTags_NR99/MP_SRF_miTags_NR99_bacteria/")

#read data 
otu_tb10_mitags_13674 <- read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/MP_SRF_miTags_NR99_bacteria.txt", head=TRUE, fill=TRUE)

#table dimensions and format before setting column names
dim(otu_tb10_mitags_13674) # 1178  125
otu_tb10_mitags_13674[1:5,1:5]

#row names = OTU name (option A)
row.names(otu_tb10_mitags_13674)<-otu_tb10_mitags_13674[,1]

#row names = row number (option B)
#rownames(otu_tb18) <- 1:nrow(otu_tb18)

otu_tb10_mitags_13674<-otu_tb10_mitags_13674[,-1]
otu_tb10_mitags_13674[is.na(otu_tb10_mitags_13674)]<-0

#column names = OTU name 
colnames(otu_tb10_mitags_13674, do.NULL = FALSE)

dim(otu_tb10_mitags_13674) #  56116   124
otu_tb10_mitags_13674[1:5,1:5]

#select 10 samples from MP-SRF miTags dataset:
tb10_mitags<- otu_tb10_mitags_13674[,c("MP0311", "MP0528", "MP0778", "MP1176", "MP1421", "MP1517", "MP1672", "MP1857", "MP2243", "MP2821")]
dim(tb10_mitags)
tb10_mitags[1:5,1:5]

#make sure that any sample has less than 13674 OTUs:
amplicons_per_sample_tb10_mitags<-colSums(tb10_mitags)
amplicons_per_sample_tb10_mitags[which(colSums(tb10_mitags)<13674)]
```

Table dimensions and content outline:

```{r starting_dataset, echo=FALSE}
dim(tb10_mitags)
tb10_mitags[1:5,1:5]
```

Minimum number of reads per station:

```{r reads_per_sample_overview1, echo=1}
min(colSums(tb10_mitags)) 
#13674
```

Maximum number of reads per station:

```{r reads_per_sample_overview2, echo=1}
max(colSums(tb10_mitags)) 
# max: 155539
```

Identification of station with higher number of reads:

```{r reads_per_sample_overview3, echo=TRUE}
amplicons_per_sample<-colSums(tb10_mitags)
amplicons_per_sample[which(colSums(tb10_mitags)>41000)]
```

Overall reads per sample:

``` {r reads_per_sample_overview4, echo=FALSE}
plot(sort(colSums(tb10_mitags)), pch=19, xlab="sample", ylab="reads per sample", cex=0.9)
```


# 2) Normalization to 13674 reads per sample

Let's normalize the original dataset by randomly subsampling 13674 reads in each station:

``` {r species_richness_rarefaction1, echo=TRUE}
library(vegan)
tb10_mitags_t<-t(tb10_mitags)
otu_tb10_mitags_t_ss13674<-rrarefy(tb10_mitags_t, 13674)
```

The normalized table shows the following dimensions and format:

```{r species_richness_rarefaction2, echo=FALSE}
dim(otu_tb10_mitags_t_ss13674)
otu_tb10_mitags_t_ss13674[1:5,1:5]
```

Its content fits with the expected normalization values (13674 reads per station):

``` {r species_richness_rarefaction3, echo=TRUE}
rowSums(otu_tb10_mitags_t_ss13674)
```

Let's check out how many OTUs don't appear in the new table:

```{r species_richness_rarefaction4, echo=1:5}
length(which(colSums(otu_tb10_mitags_t_ss13674)==0)) 
```

There are 37810 OTUs that don't show any occurrence in the normalized data. Let's remove them from the table and take a look at its final dimensions:

```{r species_richness_rarefaction5, echo=1:3}
otu_tb10_mitags_t_ss13674_no_cero<-otu_tb10_mitags_t_ss13674[,-(which(colSums(otu_tb10_mitags_t_ss13674)==0))]
dim(otu_tb10_mitags_t_ss13674_no_cero)

#The final dimensions of the normalized table are 119 1035.
#37810+6259 = 44069
```

Datasets summary:

tb10_mitags_t --> 10 1178

otu_tb10_mitags_t_ss13674_no_cero --> 10 591


# 3) General community analysis

## 3.1) Richness and evenness (Shannon index)

```{r shannon_index1, echo=FALSE}
otu_tb10_mitags_ss13674_div <- diversity(otu_tb10_mitags_t_ss13674_no_cero, index="shannon")
```

Most of the samples take Shannon Index values between 2.25 and 3.25:

```{r shannon_index2, echo=FALSE}
boxplot(otu_tb10_mitags_ss13674_div, pch=19, main="Shannon's index of diversity")
plot(sort(otu_tb10_mitags_ss13674_div), pch=19, main="Shannon's index of diversity")
```

## 3.2) Richness: OTU number

```{r richness_otu_no1, echo=FALSE}
OTUs_per_sample_10_ss13674<-specnumber(otu_tb10_mitags_t_ss13674_no_cero)
```

Lowest number of OTUs per sample:

```{r richness_otu_no2, echo=FALSE}
min(OTUs_per_sample_10_ss13674)
```

Maximum number of OTUs per sample:

```{r richness_otu_no3, echo=FALSE}
max(OTUs_per_sample_10_ss13674)
```

In most of the samples, we can identify between 200 and 230 OTUs:

```{r richness_otu_no4, echo=TRUE}
plot(sort(OTUs_per_sample_10_ss13674), pch=19)
boxplot(OTUs_per_sample_10_ss13674, pch=19)
```

## 3.3) Index of evenness

### 3.3.1) Pielou's index

```{r pielou_index_of_evenness1, echo=TRUE}
pielou_evenness_10_ss13674 <- otu_tb10_mitags_ss13674_div/log(OTUs_per_sample_10_ss13674)
```

The Pielou index (constrained between 0 and 1) takes values closer to 1 as the variation of species proportion in a sample decreases. Most of our samples get values between 0.425 and 0.625, meaning that the numerical composition of different OTUs within our samples is relatively similar:

```{r pielou_index_of_evenness2, echo=TRUE}
plot(sort(pielou_evenness_10_ss13674), pch=19)
boxplot(pielou_evenness_10_ss13674, pch=19)
```

The OTU_1, with 20053 reads, is the most abundant in the overall dataset:

```{r OTUs_overall_abundance, echo=TRUE}
head(sort(colSums(otu_tb10_mitags_t_ss13674_no_cero), decreasing=T), n=10L)
```

Most of the OTUs show few occurrences, but we don't observe the presence of a massive group of rare OTUs:

```{r OTUs_overall_abundance2, echo=TRUE}
plot(log(sort(colSums(otu_tb10_mitags_t_ss13674_no_cero), decreasing=T)), pch=19)
```

<!---
### 3.3.2) Sads
library(sads)
?sads
--->

## 3.4) Abundance Models
### 3.4.1) Rank-Abundance or Dominance/Diversity Model ("radfit")

The OTUs abundance distribution fits relativelly close to log-normal model. 

```{r radfit, echo=FALSE}
#?radfit
#otu_tb18_t[1:5,1:5]

otu_tb10_mitags_min13674_radfit<-radfit(colSums(tb10_mitags_t))
plot(otu_tb10_mitags_min13674_radfit)
```

### 3.4.2) Preston's Lognormal Model

According to Preston's lognormal model fit into species frequencies groups, we're missing ~130 species:

```{r preston_model1, echo=T}
otu_tb10_mitags_min13674_prestonfit<-prestonfit(colSums(tb10_mitags_t))
plot(otu_tb10_mitags_min13674_prestonfit, main="Pooled species")

veiledspec(otu_tb10_mitags_min13674_prestonfit)
```

When computing Preston's lognormal model fit without pooling data into groups, we seem to miss ~118 species:

```{r preston_model2, echo=4}
otu_tb10_mitags_min13674_dist_all<-prestondistr(colSums(tb10_mitags_t))
plot(otu_tb10_mitags_min13674_prestonfit, main="All malaspina")
lines(otu_tb10_mitags_min13674_dist_all, line.col="blue3")

veiledspec(otu_tb10_mitags_min13674_dist_all)
```

<!---
## 3.5) Rarefaction curve

(To be computed)

```{r rarefraction_curve, echo=TRUE}
#?rarecurve

#str(colSums(otu_tb18_t))

#otus_tb18_colsums<-colSums(otu_tb18_t)

#str(otus_tb18_colsums)
#otu_tb18_colsums<-as.matrix(otu_tb18_colsums)

#otu_tb18_colsums<-t(otu_tb18_colsums)

#otu_tb18_colsums[,1:3]

#rarecurve(otu_tb18_colsums, step = 1, 136740, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
#rarecurve(colSums(otu_tb18_t), step = 1, 136740, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
```
--->

## 3.6) Beta diversity

### 3.6.1) Dissimilarity matrix using Bray-Curtis index:

The Bray-Curtis dissimilarity, constrained between 0 (minimum distance) and 1 (highest dissimilarity) allows us to quantify the differences between samples according to the composition and relative abundance of their OTUs. In our dataset, most of the samples pairs take dissimilarity values between 0.35 and 0.55, meaning that their composition is quiet similar.

```{r beta_div1, echo=FALSE}
#?vegdist
otu_tb10_mitags_t_ss13674_no_cero.bray<-vegdist(otu_tb10_mitags_t_ss13674_no_cero, method="bray")
boxplot(otu_tb10_mitags_t_ss13674_no_cero.bray, main="Bray-Curtis dissimilarity matrix")
```

### 3.6.2) Hierarchical clustering

The only relatively evident clusters we can distinguish in the dendogram stand out in the very left side of the plot. 

(To be done: assign Longhurst provinces information to each station and check if any of the central clusters is meaningful regarding to the samples' geographical ubication)

```{r beta_div2, echo=FALSE}
#UPGMA
otu_tb10_mitags_t_ss13674_no_cero.upgma<-hclust(otu_tb10_mitags_t_ss13674_no_cero.bray, "average")
plot(otu_tb10_mitags_t_ss13674_no_cero.upgma, cex=.35, main="Samples Hierarchical Clustering")
```

### 3.6.3) Non-metric multidimensional scaling

We can identify a prominent group in the middle-left side of the NMDS and a few isolated samples in the right edge of the plot. Considering the overall distribution, we're not able to identify any informative clustering pattern. 

The stress parameter takes a value below 0.3, meaning that the plot is valid. 

```{r monoNMDS, echo=F}
#NMDS
otu_tb10_mitags_t_ss13674_no_cero.nmds<-monoMDS(otu_tb10_mitags_t_ss13674_no_cero.bray)
otu_tb10_mitags_t_ss13674_no_cero.nmds
plot(otu_tb10_mitags_t_ss13674_no_cero.nmds, main="monoMDs method")
```

When implementing a most robut function for computing NMDS plots, the result is quiet the same:

```{r metaNMDS, echo=F}
otu_tb10_mitags_t_ss13674_no_cero.meta_nmds<-metaMDS(otu_tb10_mitags_t_ss13674_no_cero.bray)
plot(otu_tb10_mitags_t_ss13674_no_cero.meta_nmds, main="metaMDS method")
```

# 4) Geographical analysis

```{r load_geo_data, echo=F, results="hide", message=F}
#load geographical ubication of stations and sort according to otu_tb10_mitags stations sequence.
MP_geo_tb10_mitags_ss13674<-read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/geo_ubication_SRF_to_1200_220sp.txt", sep="\t", header=T)

MP_geo_tb10_mitags_ss13674[,-1]
row.names(MP_geo_tb10_mitags_ss13674)<-MP_geo_tb10_mitags_ss13674[,1]


MP_geo_10sorted_ss13674<-MP_geo_tb10_mitags_ss13674[row.names(otu_tb10_mitags_t_ss13674_no_cero),]

otu_tb10_mitags_t_ss13674_no_cero[1:5,1:5]

dim(MP_geo_10sorted_ss13674)
MP_geo_10sorted_ss13674[1:5,]
otu_tb10_mitags_t_ss13674_no_cero[1:5,1:5]

#read lat-long in decimal degrees and translate into distance in km.
library(fossil)

#select only columns containing info about station, latitude and longitude.
MP_geo_10sorted_ss13674_v2<-create.lats(MP_geo_10sorted_ss13674, loc="sample", long="long", lat="lat")
head(MP_geo_10sorted_ss13674)

#create a distance matrix (lower triangle) between a list of points.
geo_distances_MP_10_ss13674<-earth.dist(MP_geo_10sorted_ss13674_v2, dist = TRUE)
head(geo_distances_MP_10_ss13674)
dim(geo_distances_MP_10_ss13674)

geo_distances_MP_10_ss13674<-as.matrix(geo_distances_MP_10_ss13674)
dim(geo_distances_MP_10_ss13674)

#geo distances dataset ready to use "geo_distances_MP_euks"
```

Working datasets:

1) Community matrix: otu_tb10_mitags_t_ss13674_no_cero

```{r working_datasets1, echo=T}
dim(otu_tb10_mitags_t_ss13674_no_cero)
otu_tb10_mitags_t_ss13674_no_cero[1:5, 1:5]
```

2) Community Bray-Curtis: otu_tb10_mitags_t_ss13674_no_cero.bray

```{r working_datasets2, echo=2}
#dim(otu_tb10_mitags_t_ss13674_no_cero.bray)
otu_tb10_mitags_t_ss13674_no_cero.bray<-as.matrix(otu_tb10_mitags_t_ss13674_no_cero.bray)
```

3) Stations distances in km: geo_distances_MP_10_ss13674

```{r working_datasets3, echo=T}
dim(geo_distances_MP_10_ss13674)
```

Communities quickly change their composition across geographical distances:

```{r working_datasets4, echo=T}
plot(geo_distances_MP_10_ss13674, otu_tb10_mitags_t_ss13674_no_cero.bray, pch=19, cex=0.4, xlab="Geopgraphical distances", ylab="Bray-Curtis dissimilarities")
```

## 4.1) Mantel correlograms

When implementing the Mantel Test, we obtain a very low statistic, meaning that the correlation between samples dissimilarity and geographical distances is weak; besides, it is not significant, so we shouldn't consider the test meaningful.

```{r mantel_correlogram1, echo=T}
mantel(geo_distances_MP_10_ss13674, otu_tb10_mitags_t_ss13674_no_cero.bray)
```

Maximum distance between samples:

```{r mantel_correlogram2, echo=F}
max(geo_distances_MP_10_ss13674)
```

Minimum distance between samples:

```{r mantel_correlogram3, echo=F}
min(geo_distances_MP_10_ss13674)
```


<!---
Correlograms:

```{r mantel_correlogram4, echo=T}
MP_10_ss13674_mantel_correl_by_1000km<-mantel.correlog(otu_tb10_mitags_t_ss13674_no_cero.bray, D.geo=geo_distances_MP_10_ss13674, break.pts=seq(0,20000, by=1000))
plot(MP_10_ss13674_mantel_correl_by_1000km)

#MP_10_ss13674_mantel_correl_by_100km<-mantel.correlog(otu_tb10_mitags_t_ss13674_no_cero.bray, D.geo=geo_distances_MP_10_ss13674, break.pts=seq(0,20000, by=100))
#plot(MP_10_ss13674_mantel_correl_by_100km)
```
--->



# 5) Abundance & occurence

```{r OTUs_mean_relative_abund, echo=F, results="hide"}
otu_tb10_mitags_t_ss13674_no_cero[1:5,1:5]
otu_tb10_mitags_t_ss13674_no_cero_t<-t(otu_tb10_mitags_t_ss13674_no_cero)

colSums(otu_tb10_mitags_t_ss13674_no_cero_t)

#local abundance percentage
otu_tb10_mitags_t_ss13674_no_cero_t.rabund<-otu_tb10_mitags_t_ss13674_no_cero_t/13674

colSums(otu_tb10_mitags_t_ss13674_no_cero_t.rabund)
otu_tb10_mitags_t_ss13674_no_cero_t.rabund[1:5,1:5]

#OTUs mean relative abundance
otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means<-rowMeans(otu_tb10_mitags_t_ss13674_no_cero_t.rabund) 
otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means<-as.data.frame(otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means)

head(otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means)
```

```{r OTUs_occurence, echo=F, results='hide'}
otu_tb10_mitags_t_ss13674_no_cero_t.rabund.occur<-otu_tb10_mitags_t_ss13674_no_cero_t.rabund
otu_tb10_mitags_t_ss13674_no_cero_t.rabund.occur[otu_tb10_mitags_t_ss13674_no_cero_t.rabund.occur>0]<-1
otu_tb10_mitags_t_ss13674_no_cero_t.rabund.occur[1:5,1:5] ### presence - absence table

#percentage of occurence in overall stations
otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means.occurence_perc<-as.data.frame(100*(rowSums(otu_tb10_mitags_t_ss13674_no_cero_t.rabund.occur)/10))

str(otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means.occurence_perc)
```

```{r merge_rabund_peroccur, echo=F, results='hide'}
otu_tb10_mitags_ss13674_rabund_percoccur<-merge(otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means,otu_tb10_mitags_t_ss13674_no_cero_t.rabund_means.occurence_perc, by="row.names")

colnames(otu_tb10_mitags_ss13674_rabund_percoccur)<-c("OTUs","mean_rabund","perc_occur")
otu_tb10_mitags_ss13674_rabund_percoccur[1:5,]

row.names(otu_tb10_mitags_ss13674_rabund_percoccur)<-otu_tb10_mitags_ss13674_rabund_percoccur[,1]
otu_tb10_mitags_ss13674_rabund_percoccur<-otu_tb10_mitags_ss13674_rabund_percoccur[,-1]
otu_tb10_mitags_ss13674_rabund_percoccur[1:5,]
```

In the following plot, we can appreciate the OTUs distribution according to their percentage of occurence and relative abundance. The red line keeps up OTUs that occur in more than 80% of the samples, the green line limits regionally rare OTUs (< 0.001%), and the blue one restricts regionally abundant OTUs (> 0.1%).

```{r abund_vs_occurence_table, echo=F}
plot(otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund,otu_tb10_mitags_ss13674_rabund_percoccur$perc_occur, log="x", pch=19, cex=0.8, xlab="Mean relative abundance", ylab="Percentage of occurence")
abline(h=80, col="red") #occurence higher than 80%
abline(v=0.00001, col="green") #rare OTUs
abline(v=0.001, col="blue") #cosmopolitan OTUs

#Conventional limits:
#Regionally rare     = 0.00001
#Regionally abundant = 0.001
```

Regionally abundant OTUs (relative abundance over 0.1%):

```{r abundant_OTUs, echo=7}
#regionally abundant
tb10_mitags_ss13674_abundant<-otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund > 0.001,]

tb10_mitags_ss13674_abundant_sorted<-tb10_mitags_ss13674_abundant[order(tb10_mitags_ss13674_abundant$mean_rabund, tb10_mitags_ss13674_abundant$perc_occur, decreasing = T), c(1,2)]

tb10_mitags_ss13674_abundant_sorted
dim(tb10_mitags_ss13674_abundant_sorted)
```

Proportion of regionally abundant OTUs (%):

```{r abundant_OTUs2, echo=F}
#there are 84 regionally abundant OTUs.
(184/6259)*100 # = 2.94% of the OTUs are regionally abundant

#length(row.names(otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund > 0.001,])) # 83 OTUs
#row.names(otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund > 0.001,])
```

Cosmopolitan OTUs (abundance over 0.1% and occurence in more than 80% of samples):

```{r select_cosmopolitan, echo=6}
otu_tb10_mitags_ss13674_rabund_cosm<-otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund > 0.001,]
otu_tb10_mitags_ss13674_rabund_poccur_cosm<-otu_tb10_mitags_ss13674_rabund_cosm[otu_tb10_mitags_ss13674_rabund_cosm$perc_occur > 80,]
otu_tb10_mitags_ss13674_cosmop_sorted<-otu_tb10_mitags_ss13674_rabund_poccur_cosm[order(otu_tb10_mitags_ss13674_rabund_poccur_cosm$perc_occur, otu_tb10_mitags_ss13674_rabund_poccur_cosm$mean_rabund, decreasing = T), c(1,2)]

otu_tb10_mitags_ss13674_cosmop_sorted
dim(otu_tb10_mitags_ss13674_cosmop_sorted)
```

Proportion of cosmopolitan OTUs (%):

```{r percentage_cosmopolitan, echo=F}
(168/6259)*100
```

Number and proportion (%) of rare OTUs:

```{r rare_OTUs, echo=1}
dim(otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund < 0.00001 & otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund >0,])
 
(2050/6259)*100 # = 32.75% of the OTUs are regionally rare
```

<!---
```{r otu_col_chech, echo = T}
dim(otu_tb10_mitags_ss13674_rabund_percoccur)
dim(tb10_mitags)
tb10_mitags[1:5,1:5]

#rare OTUs:
# length(row.names(otu_tb10_mitags_ss13674_rabund_percoccur[otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund < 0.00001 & otu_tb10_mitags_ss13674_rabund_percoccur$mean_rabund >0 ,])) # 325 OTUs
```
--->